---
title: "qPCR template for Calculating primer efficiencies"
author: "Adrija Kalvisa"
date: "28/01/2020"
output: 
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Experimental setup:
Experiment ID:  
Date of running qPCR:  
Primers:  
Master mix composition:  
qPCR setup:  

### About this analysis

The sample loading schemes usually are in a "wide" format (imitating the 96 or 384 well PCR plate layout) while the Cp values are obtained in a "long" format. In order to avoid time-consuming and error-prone manual conversion from one format to another, this script provides a framework to automatize merging the Calculated Cp values with sample IDs.

## Step 1: Calculate primer efficiencies

This part you have to change:

```{r}
#pathToSampleLayout = "qPCR SMARCE1 Molm13.xlsx"
#pathToCpValues = "SMARCE1_Lightcycler480_output.txt"

# if the file is in the same folder as this script then you can provide only the sample name. 
pathToSampleLayout = "12Sept2019-validation primers test.xlsx"
pathToCpValues = "primer_validation_Cp_values.txt"

referencePrimers="B2M" 
```

The rest of the script should work without your supervision. I suggest, however, double checking that the right bits are asigned to "primer", "treatment", "biological_replicate", "technical_replicate" and "primer_ratio" columns. This bit can be modifoed in 'Load Sample Layout' section. 

### Libraries
```{r, message=FALSE, results='hide'}
library(readxl)

#install.packages("platetools")
library(platetools) 

#install.packages("pcr")
library(pcr)

# install.packages(“ggplot2”)
library(ggplot2)

# install.packages("viridis")
library(viridis)

# install.packages(“reshape2”)
library(reshape2)   

# install.packages("dplyr")
library(dplyr)

# install.packages("tidyr")
library(tidyr)
```

### Custom-made functions

`entry2columns(my_column, list_of_informative_cathegories, separator_string)`  
separates sample information into relevant user-specified cathegories ("primer", "treatment", "biological_replicate", "technical_replicate", "primer_ratio").

```{r entry2columns, echo=FALSE}
entry2columns = function(my_column, list_of_informative_cathegories, separator_string){

  # this line splits the "my_column" into n columns of selected cathegories. n = number of selected cathegories, 1-5
  columns = 
    stringr::str_split_fixed(my_column, separator_string, length(list_of_informative_cathegories))
  
  # rename columns so that they are consistent with downstream analysis
  colnames(columns) = list_of_informative_cathegories
    
  return(columns)
  
}
```

`transformWellNumbers(posString)` turns numbers "A1"" "B2" "C3" to "A01" "B02" "C03".
```{r transformWellNumbers, echo=FALSE}
transformWellNumbers = function(posString){
  # Transforms A1 to A01 B1 to B01 (works for numbers < 10)
  return(gsub("(?<![0-9])([0-9])(?![0-9])", "0\\1", posString, perl = TRUE)
  )
}
```

`readFromFile_CpValues(pathToFile)` reads the output .txt file from Lightcycler480 machine.
```{r readFromFile_CpValues, echo=FALSE}
readFromFile_CpValues = function(pathToFile){
  # read the data from the given file path, assuming tab delimiter and comma as decimal separator. 
  # the Well IDs should be stored in column "Pos", sample names in column "Name", Cp values in "Cp"
  
data = 
  data.table::fread(
    pathToFile, 
    na.strings=NULL , 
    dec=",", 
    sep="\t", 
    skip=1, 
    select = c("Pos", "Name" ,"Cp"), 
    data.table=FALSE
    )

  data$Pos = 
  transformWellNumbers(data$Pos)

  return(data)
}
```

`readFromFile_SampleLayout(pathToFile)` reads a specifically formatted text file that contains sample layout information. 
```{r readFromFile_SampleLayout, echo=FALSE}
readFromFile_SampleLayout = function(pathToFile) {
  
  #load data from .txt file path
  #tmp = 
  #  read.delim(pathToFile,row.names = c("A", "B", "C", "D", "E", "F", "G", "H"))

  # load data from .xls file path
  output = 
    read_excel(pathToFile, range = "B2:M10")
  
  row.names(output) = c("A", "B", "C", "D", "E", "F", "G", "H")
    
  # drop the first column containing rownames
  #output = 
  #  tmp[,-1]
  
  # rename the column names
  colnames(output) = 
    as.character(c(1:12))
  
  return(output)
}
```

`plate2table(dataFrame)`
transforms the sample layout from "wide" to "long" format.
```{r plate2table, echo=FALSE}
plate2table = function(wideDataFrame){
  
  # make the files into long format
  long = melt(as.matrix(wideDataFrame))
  
  # add the position names, making sure they are 0-padded. this is important for platetools package
  long$Pos = transformWellNumbers(paste0(long$Var1, long$Var2))
  
  return(long[,c("Pos", "value")])
}
```

`meanWithoutNA` calculates means of Cp when some values are missing
```{r meanWithoutNA, echo=FALSE}
meanWithoutNA = function(x){
  mean(x, na.rm=TRUE)
  }

countNA = function(x){
  sum(is.na(x))
  }
```

`ratio2conc("characterString")` turns primer ratio ("1:100") into relative primer concentration ("0.01")
```{r ratio2conc, echo=FALSE}
ratio2conc = function(ratioColumn){
 as.numeric(stringr::str_split_fixed(ratioColumn, ":", 2)[,2])^-1
}
```


### Load sample Layout

```{r}
# load sample layout
primer_validation_sample_layout = readFromFile_SampleLayout(pathToSampleLayout)

# turn sample layout into "long" list format
sample_list = plate2table(primer_validation_sample_layout)

# split the loading information into separate columns and add back to the long table
sample_list = sample_list$value %>% entry2columns(c("primer", "primer_ratio"), " ") %>% cbind(sample_list)
```

`entry2columns(c("primer", "primer_ratio"), " ")` splits the column into two new columns by splitting the column where the space `"" "` is. Note that if you have several different delimiters, you can separate them with a `|` which means `OR`. For example, `"-| "` will be used to split "RNF141-1 1:100" into "RNF141" "1" "1:100".

other examples for: 
```{r}
# ideally the column names should be some of the following:
# "primer", "treatment", "biological_replicate", "technical_replicate", "primer_ratio"

"PrimerABC_treatmentDEF_repGHI" %>% entry2columns(c("primer", "treatment", "biological_replicate"), "_")
"PrimerABC-treatmentDEF repGHI" %>% entry2columns(c("primer", "treatment", "biological_replicate"), "-| ")
"PrimerABC-treatmentDEF repGHI 1:100" %>% entry2columns(c("primer", "treatment", "biological_replicate", "primer_ratio"), "-| ")
```


```{r}
# make sure that each well has an unique naming - I assume that identical come from "technical replicates"
sample_list = 
  sample_list %>%
  group_by(primer, primer_ratio) %>%    
  mutate(technical_replicate = row_number()) %>% 
  select(-"value")
```

`group_by(primer, primer_ratio)` finds the unique combinations of primer+primer_ratio   
`mutate(technical_replicate = row_number())` counts how many times these unique values were encountered and saves these counts in column named 'technical replicate'
`select(-"value")` drops the column named 'value' from  sample_list because it has become redundant

how does sample list look like?
```{r}
sample_list
```

show the Cp values in the plate layout
```{r}
# load Cp values
primer_validation_Cp_values = readFromFile_CpValues(pathToCpValues)

# show the plate layout
platetools::raw_map(
  data = primer_validation_Cp_values$Cp,
  well = primer_validation_Cp_values$Pos) + 
  ggtitle("primer validation example") +
  theme_light() +
  scale_fill_viridis()
```

```{r}
# Merge Cp values with sample information
Samples_Cp_long = 
  merge(sample_list, primer_validation_Cp_values, by="Pos")

# Extract dilution factor
Samples_Cp_long$primer_conc = ratio2conc(Samples_Cp_long$primer_ratio)
```

what primers are tested?
```{r}
Samples_Cp_long$primer %>% as.character() %>% unique()
```

Make a data frame where each column contains one primer set and each row contains one sample (1 sample = 1 eppendorf tube)

```{r}
data_for_pcr_table = 
  Samples_Cp_long[Samples_Cp_long$primer != referencePrimers, c("primer","technical_replicate","primer_conc","Cp")]

pcr_table = 
  tidyr::pivot_wider(
  data = data_for_pcr_table,
  names_from = primer,
  values_from = Cp,
  values_fn = list(Cp = meanWithoutNA)
  ) %>% 
  select(-"technical_replicate")
```

how does pcr table look like?
```{r}
pcr_table
```

Calculate and plot the efficacy using `qpcR::pcr_assess()` function
```{r}
pcr_assess(
  df = pcr_table[,-1],
  amount = pcr_table$primer_conc,
  method = 'standard_curve')

# plot the slope
pcr_assess(
 df = pcr_table[,-1],
 amount = pcr_table$primer_conc,
 method = 'standard_curve', 
 plot=TRUE)
```

More analysis tips: 
https://cran.r-project.org/web/packages/pcr/vignettes/qpcr_analysis.html

## Further analysis
More guidelines on analysing PCR results can be found in 
[pcr package vignette](https://cran.r-project.org/web/packages/pcr/vignettes/qpcr_analysis.html)
